AppPAL for Android

### 2015-04-01 paper.tex
wrote most of details

\documentclass[a4paper]{article}
\usepackage{defaultstyle}
\usepackage{localstyle}

\newcommand{\comment}[1]{\textsf{#1}}

\title{Bringing AppPAL to Android}
%\subtitle{Capturing and Checking App Installation Policies}
\author{Joseph Hallett \and David Aspinall}
\date{\today}

@@

\begin{abstract}
  The abstract goes\ldots
  \begin{itemize}
    \item \comment{Sentence 1: State the problem}
    \item \comment{Sentence 2: Why is it an interesting problem}
    \item \comment{Sentence 3: What does my solution achieve}
    \item \comment{Sentence 4: What follows from my solution}
  \end{itemize}
\end{abstract}

\section{Introduction \comment{1 page}}

\section{The Problem \comment{1 page}}

\section{My idea \comment{2 pages}}

\section{The details \comment{5 pages}}

>>>>

% vim: set nowrap tw=0:
\documentclass[a4paper]{scrartcl}
\usepackage{defaultstyle}
\usepackage{localstyle}

\newcommand{\comment}[1]{}%\textsf{#1}}

\title{AppPAL for Android}
\subtitle{Capturing and Checking App Installation Policies}
\author{Joseph Hallett \and David Aspinall}
\date{\today}

@@

\begin{abstract}
  % \begin{itemize}
  %   \item \comment{Sentence 1: State the problem}
  When a user wants to pick an app on Android they are forced to judge apps based only on the information presented to them by the store.
  %   \item \comment{Sentence 2: Why is it an interesting problem}
  Employers must rely on employees using company devices properly to enforce security policies about apps at work.
  Whilst some users spend time to pick apps with care others don't and become frustrated at apps that steal personal information.
  %   \item \comment{Sentence 3: What does my solution achieve}
  We present AppPAL: a policy language that can enforce app installation policies using statements from trusted third parties and delegation relationships.
  %   \item \comment{Sentence 4: What follows from my solution}
  This gives us a rigorous and flexible framework for enforcing and comparing and app policies and the trust relationships surrounding them.
  % \end{itemize}
\end{abstract}

\section{Introduction \comment{1 page}}
\label{sec:introduction}

%\comment{The problem is\ldots}
How do you pick which apps to install on your phone?  
App stores give some information about their apps; such as permissions and review scores.
Ultimately the decision which apps to use must be made by the user, however.

Not all apps are suitable. 
There is a large amount of \ac{pus} available for Android~\citep{Truong:2014bi,Svajcer:2013tp}. 
Employees are increasingly using their own phones for work (the \ac{byod} scheme.)
Their company may want to restrict which apps their employees can use.
The \ac{ciso} or IT department may set a policy to prevent information leaks.
Some users worry apps will misuse their personal data.
They avoid apps which can access their location, or contact data.
They apply their own personal security policy when downloading apps.

Currently these policies are enforced by hand.
This is error-prone and can be inconsistent.
A better solution is to use a logic of authorization.
The policy can be written in the logic and enforced automatically.

%\comment{My contribution is\ldots}
We present AppPAL, an authorization logic for reasoning about Android apps.
The language is an instantiation of SecPAL~\citep{Becker:2006vh} tailored to making statements about apps.
It allows us to reason about apps using statements from third parties.
We can express trust relationships in these speakers and use constraints to perform additional checks at evaluation time.
This allows us to enforce deeper and more complex policies than existing policy tools like Kirin~\citep{Enck:2009ko}.

\begin{itemize}
  %\item \comment{Specifically I have}
  \item Described several scenarios where an authorization logic could replace informal policies for companies and users (\autoref{sec:problem}.) 
    \comment{Provide forward refs saying where we do this and about what}
  \item Shown how app installation policies can be expressed in AppPAL
    including how a corporate \ac{byod} policy could be written, and how a user's personal security policy could be formalized (\autoref{sec:idea}.)
  \item Implemented the AppPAL language.  
    Our implementation runs on Android and the JVM (\autoref{sec:details}.)
\end{itemize}

\section{The Problem\comment{1 page}}
\label{sec:problem}

Suppose \emph{Alice} works for \emph{Bob}.
Bob allows Alice to use her personal phone as a work phone, but he has some specific concerns.
\begin{itemize}
  \item Alice shouldn't have any apps that can track her movements.
    The testing labs are at a secret location and it mustn't be leaked.
  \item Apps should come from a reputable source, such as the Play store.
  \item Bob uses an \ac{av} program by McAfee. 
    It should check all apps before they're installed.
\end{itemize}

To ensure this policy is fully met Alice could promise to follow it; but this is error-prone---what if she makes a mistake or misses something?
Bob's policy could be implemented using existing tools.
\emph{Google's Device Policy for Android}\footnote{\url{https://play.google.com/store/apps/details?id=com.google.android.apps.enterprise.dmagent}}
  could configure Alice's device to disallow apps from outside the Play Store.
Various tools~\citep{Backes:2012vm,Jeon:2012ki,Hornyack:2011wq} can control the permissions or data an app can get.
These could be used to used to ensure no location data is ever obtained.
Alternately other tools~\citep{Enck:2009ko,Fritz:2013vi,Gordon:2015et} could check that the locations are ever leaked to the web.
Various anti-virus programs are available for Android---one could be installed on Alice's phone.

Whilst we \emph{could} implement Bob's policy using existing tools, it isn't a great solution.
It isn't flexible, Alice needs to alter and remove the software on her phone if Bob changes his policy or she changes jobs.
It isn't clear what an app must do to be installed, or what checks must have been done if it already has been put on the phone.
The link between Alice, the user; Bob, the policy setter; and the tools Bob trusts to implement his policy isn't immediately apparent.

This last point is particularly pertinent with respect to \textsc{CVE-2015-2077}.
In 2015 Lenovo was found to be shipping laptops with the \emph{Superfish} malware pre-installed.
Lenovo had also installed an \ac{av} package on their laptops; this \ac{av} did recognise Superfish as malware.
Unfortunately they had configured it to ignore Superfish\footnote{The malware was an ad framework designed to show users products similar to those they viewed on the web.  Unfortunately it did so by man-in-the-middling all SSL traffic using a shared private key.}.

From a user's perspective there was a delegation of trust to Lenovo to detect malware.
Lenovo then delegated further to McAfee to supply the antivirus; but left an exception that they would allow Superfish.
A user might wonder where the breach of trust occurred and who is to blame?  
Is it the \ac{av} failing to spot the malware? 
Has someone else configured something incorrectly?
How should they fix the problem?

If we write this in the AppPAL policy language the cause of the breach of trust becomes apparent; as does the fix.
\begin{lstlisting}[numbers=left, escapeinside={@}{@}, label={lst:lenovo}]
"user" says "lenovo" can-say inf File isSafe. @\label{lst:lenovo_trust}@
"lenovo" says "mcafee" can-say inf File isSafe.
"lenovo" says "C:\System\superfish" isSafe. @\label{lst:lenovo_unsafe}@
\end{lstlisting}
The fault lies with \autoref{lst:lenovo_unsafe}.  
Lenovo has caused the problem (they \emph{said} it was safe).
The fix is to revoke this statement, or to revoke \autoref{lst:lenovo_trust} and find a different \ac{av} supplier.

It is important to stress that we are not trying to solve the problem of detecting malware.
There are many tools, and much research, that already does this.
Our problem is different. 
Given there are some apps you want and others you don't:
  how can you express your preferences in such a way that they can be enforced automatically.
Further more how can we show the trust relationships used to make these decisions clearly and precisely.

\section{My idea \comment{2 pages}}
\label{sec:idea}

Returning to Alice and Bob's dilemma in \autoref{sec:problem}, we propose a solution.
Instead of using several different tools to enforce Bob's policy disjointedly, we use an authorization logic.
In \autoref{lst:corporate} we give an AppPAL policy implementing Bob's app concerns on Alice's phone.
A pictorial representation, is given in \autoref{fig:bobs_policy}; color is used to split statements by each speaker.

\begin{figure}
  \includegraphics{figures/bob-policy.eps}
  \caption{Pictoral representation of Alice's app install policy.}
\label{fig:bobs_policy}
\end{figure}

AppPAL is an instantiation of SecPAL~\citep{Becker:2006vh} for app installation policies. 
SecPAL is a logic of authorization for access control decisions in distributed systems.
It has a clear and readable syntax, as well as rich mechanisms for delegation and constraints.
SecPAL has already been used as a basis for other policy languages in areas such as privacy preferences~\citep{Becker:2009ula} and data-sharing~\citep{Aziz:2011vt}.
We present AppPAL as a new version of SecPAL, targeting apps on mobile devices.

\begin{lstlisting}[label={lst:corporate}, 
                   caption={An AppPAL policy implementing Bob's security requirements},
                   float,
                   numbers=left,
                   escapeinside={@}{@}]
"alice" says "bob" can-say inf App isInstallable. @\label{lst:corporate_1}@
"bob" says App isInstallable @\label{lst:corporate_2}@
  if "no-tracking-policy" isMetBy(App),
     "reputable-source-policy" isMetBy(App),
     "anti-virus-policy" isMetBy(App).
"bob" says "reputable-source-policy" isMetBy(App) @\label{lst:corporate_3}@
  if App isBuyable.
"bob" says "google-play" can-say 0 App isBuyable. @\label{lst:corporate_4}@
"bob" says "anti-virus-policy" isMetBy(App) @\label{lst:corporate_5}@
  if App isAnApp
  where
    mcAfeeVirusCheck(App) = False.
"bob" says "no-location-permissions" can-act-as "no-tracking-policy". @\label{lst:corporate_6}@
"bob" says "no-location-permissions" isMetBy(App) @\label{lst:corporate_7}@
  if App isAnApp
  where
    hasPermission(App, "ACCESS_FINE_LOCATION") = False,
    hasPermission(App, "ACCESS_COURSE_LOCATION") = False.
\end{lstlisting}

In \autoref{lst:corporate_1} Alice gives Bob the ability to specify whether an \code{App} (a variable) \code{isInstallable} (a predicate).
She allows him to delegate the decision further if he choses (\code{can-say inf}).
Next in \autoref{lst:corporate_2} Bob specifies his concerns as policies to be met (the \code{isMetBy()} predicate that takes an app as its argument).
If Bob can be convinced all these policies are met then he will say the \code{App isInstallable}.
In \autoref{lst:corporate_3} and \autoref{lst:corporate_4} Bob specifies that an app meets the \code{reputable-source-policy} if the \code{App isBuyable};
  with \code{"google-play"} specified as the decider of what is buyable or not.
This time Google is not allowed to delegate the decision further (\code{can-say 0}).
In other words Google is not allowed to specify Amazon as a supplier of apps as well.
Google must say what is buyable directly for Bob to accept it.
Bob specifies the \code{"anti-virus-policy"} in \autoref{lst:corporate_5}.
Here we use a constraint.
When checking the policy the \code{mcAfeeVirusCheck} should be run on the \code{App}.
Only if this returns \code{False} will the policy be met.
To specify the \code{"no-tracking-policy"} Bob says that the \code{"no-location-permissions"} rules implement the \code{"no-tracking-policy"} (\autoref{lst:corporate_6}).
Bob specifies this in \autoref{lst:corporate_7} by checking the app is missing two permissions.

Say Alice wants to install a new app (\code{com.facebook.katana}) on her phone.
To meet Bob's policy the AppPAL policy checker needs to collect statements to show the app meets the \code{isInstallable} predicate.
Specifically it needs:
\begin{description}
  \item[\texttt{"bob" says "com.facebook.katana" isAnApp.}] 
    A simple typing statement that can be generated for all apps as they are encountered.
  \item[\texttt{"google-play" says "com.facebook.katana" isBuyable.}] 
    Required to convince Bob the app came from a reputable source. 
    It should be able to obtain this statement from the Play store as the app is available there.
  \item[\texttt{"bob" says "anti-virus-policy" isMetBy("com.facebook.katana").}]
    She can obtain this by running the \ac{av} program on her app.
  \item[\texttt{"bob" says "no-locations-permissions" isMetBy("com.facebook.katana").}]
    Needed to show the App meets Bob's no-tracking-policy.
    Bob will say this if after examining the app the location permissions are missing.
\end{description}
These last two statements require the checker to do some extra checks to satisfy the constraints.
To get the third statement it must run the \ac{av} program on her app and check the result.
The results from the \ac{av} program may change with time as it's signatures are updated.  
Consequently the checker must re-run this check every time it wants to obtain the statement connected to the constraint.
For the forth statement the checker needs to check the permissions of the app.
It does this by looking in the \texttt{MANIFEST.xml} inside the app itself.

In this scenario we have imagined Alice wanting to check the apps as she installs them.
Alternatively we could imagine Bob wanting a personalised app store where all apps sold meet his policy.
With AppPAL this can be implemented by taking an existing store and selectively offering only the apps which meet the policy.

\section{The details \comment{5 pages}}
\label{sec:details}

AppPAL is implemented as a library for Android. 
A standalone Java version also exists for testing.  
The parser is implemented using ANTLR4.
Code, tests, and prebuilt APKs are available from Github\footnote{\url{https://github.com/bogwonch/AppPAL}}.

\subsection{The Language}

The structure of an AppPAL assertion is shown in \autoref{fig:assertion}.

\begin{figure}
  \centering
  \begin{equation*}\small
    \begin{array}{r l}
      \overbrace{\text{\ttfamily "user"}}^{\text{speaker}} & 
      \text{\ttfamily says }\overbrace{\overbrace{\text{\ttfamily App }}^{\text{subject}}\overbrace{\text{\ttfamily isInstallable}}^{\text{predicate}}}^{\text{fact}} \\
      & \overbrace{\text{\ttfamily if App isFree}}^{\text{condition}} \\
      & \overbrace{\text{\ttfamily where hasPermission(App, "INTERNET") = True}}^{\text{constraint}}. 
    \end{array}
  \end{equation*}
  \caption{Structure of an AppPAL assertion.}
\label{fig:assertion}
\end{figure}

In the \citeauthor{Becker:2006vh}'s paper~\citep{Becker:2006vh} they leave the choice of predicates, and constraints for their SecPAL open.
With AppPAL we make explicit our predicates and how they relate to Android.  
Some of these predicates require arguments.
We add a light typing system to help ensure these predicates have the correct arguments.

Specifically we define the following predicates:
\begin{description}
  \item[\texttt{App isAnApp}, \texttt{Policy isAPolicy}\ldots] 
    Simple typing relations.  
    The app \text{App} is an App, and so on.

  \item[\texttt{App isInstallable}]
    Used to indicate that an App meets the install policy for the device.
    Showing an app satisfies this predicate is usually the goal of evaluating AppPAL.

  \item[\texttt{Policy isMetBy(App)}]
    Policies let you split app behaviour into sub-policies.
    For example in \autoref{sec:idea} we showed how Bob's installation policy could be written using three \texttt{isMetBy} statements.
    Splitting policies allows greater control of how each one is checked.
    We can delegate checking a policy to an expert using the \code{can-say} statement.
    We can specify more detailed checks using the \code{can-act-as} statements.
\end{description}

We also add the following constraint functions which check simple properties of the app:
In every case the check can be done by examining the app itself.
\begin{description}
  \item[\texttt{hasPermission(App, String)} : Bool]
    Checks the permissions requested in the app's manifest to see if the string is included in them.

  \item[\texttt{hasPackage(App, String)} : Bool]
    Checks the app to see if a package called \emph{String} is included within it.
    Examples might include aggressive advertisement frameworks a user wishes to ban.

  \item[\texttt{hasNativeCode(App)} : Bool]
    Checks to see whether the app contains native code.
    Native (i.e. non-Dalvik) code is potentially more dangerous than the sandboxed Dalvik versions.
    It is typically used for performance sensitive code (i.e. games).

  \item[\texttt{isSignedBy(App, Speaker)} : Bool]
    Checks the \texttt{CERT.RSA} used to sign the App and compares it with the certificate used to identify a speaker.

  \item[\texttt{hasVersion(App)} : Int]
    Gets the version number of an app.
\end{description}

We also add some constraints that do more complex checks:
\begin{description}
  \item[\texttt{googlePlayRating(App)} : Int]
    Used to get the review score (out of 5) from the Google Play store.
    This could be extended to other reviewers.

  \item[tool\texttt{Check(App, Config)} : Bool]
    Uses the \emph{tool} to run a check on an app.
    This allows static analysis tools to run inferences on apps on demand.
    Static analysis can be expensive for devices to perform.
    An alternative would be to query a database of results~\citep{Noel:2014vt}.

  \item[tool\texttt{CheckEvidence(App, Config, Evidence)} : Bool]
    Digital evidence and proof~carrying~code~\citep{Necula:1996tr,Stark:2009uc} allow you to check an analysis result without repeating it.
    In a mobile phone this can be preferable to doing the inference as draining batteries to perform checks is undesirable.
    Tools, such as \emph{evicheck}~\citep{Noel:2014vt}, can produce digital evidence for Android apps.
    This constraint function allows us to incorporate these tools into AppPAL policies.
\end{description}

\subsection{Evaluation}

To evaluate AppPAL we implement the SecPAL evaluation rules shown in \autoref{fig:rules}.
We do not use the DatalogC~\citep{Li:2003ix}~based translation and evaluation algorithm suggested by \citeauthor{Becker:2006vh}.
Rather we implement the rules directly in Java. 
Brief pseudocode is given in \autoref{fig:pseudocode}
Like \citeauthor{Becker:2006vh} we make use of an assertion context to store known statements.
We also store intermediate results to avoid re-computation.
On a mobile device memory is at a premium.
We would like to keep the context as small as possible.
To do this we allow some assertions to be imported at evaluation-time.

\begin{figure}
\[\footnotesize
% AC, D ⊧ A says B can-act-as C   AC, D ⊧ A says C verbphrase 
% ----------------------------------------------------------- (can-act-as)
%                 AC, D ⊧ A says B verbphrase
\infer[\text{\tiny can-act-as}]{
  AC, D \models \text{\tt A says B verbphrase}}{%
  AC, D \models \text{\tt A says B can-act-as C} & %
  AC, D \models \text{\tt A says C verbphrase}} 
\]
\[\footnotesize
% AC, ∞ ⊧ A says B can-say D fact   AC, D ⊧ B says fact
% ----------------------------------------------------- (can-say)
%                   AC, ∞ ⊧ A says fact
\infer[\text{\tiny can-say}]{
  AC, \infty \models \text{\tt A says fact}}{%
  AC, \infty \models \text{\tt A say B can-say $D$ fact} & %
  AC, D \models \text{\tt B says fact}} 
\]
\[\footnotesize
% (A says fact if fact₁, …, factₖ, c) ∈ AC
% AC, D ⊧ A says factᵢθ  ∀i ∈ {1…k}    ⊧cθ    vars(factθ) = ∅
% ----------------------------------------------------------- (cond)
%                    AC, D ⊧ A says factθ
\infer[\text{\tiny cond}]{
  AC, D \models \text{\tt A says fact$\theta$}}{
  \begin{array}{c}
    (\text{\tt A says fact if fact$_1$, \ldots, fact$_k$ where c}) \in AC \\
    \forall i \in 1\cdots k.\;AC, D \models \text{\tt A says fact$_i\theta$}
  \end{array} &
  \models \mathtt{c}\theta &
  \text{\sl vars}(\text{\tt fact$\theta$})}
\]
\caption{AppPAL's evaluation rules.}
\label{fig:rules}
\end{figure}

This gives us greater control of the evaluation and how the assertion context is created.
For example, when checking the \code{isAnApp} predicate; 
  we can fetch the assertion that the subject is an app based on the app in question.
Similarly when we use a statement from \emph{Bob} that \emph{Google-Play can-say} whether an app is purchasable;
  it is sensible to go fetch from the store whether the app is salable and make Google say it then and there.

In SecPAL there is no notion of type.
All entities are either constants (effectively strings) or variables.
There is no differentiation made between a \emph{voiced} constant (one who speaks),
and a \emph{subject} (one that is the subject of a verb-phrase, or predicate argument).
In AppPAL the subjects are usually apps or policies.
Apps do not utter facts.
Consequently there is no need to check them as potential speakers of \emph{can-say} statements.
Similarly if a constant never appears as a subject of a statement, as most speakers don't,
  there is no need to check it as a possible subject for a \emph{can-act-as} statement.
We pre-process the assertion context to find \emph{voiced} and \emph{subject} constants.
This allows us to cut the time spent searching for proofs.

\begin{figure}
\begin{minipage}[b]{0.5\linewidth}
\begin{lstlisting}[language=Ruby, basicstyle=\ttfamily\scriptsize, keywordstyle=\scriptsize\slshape, columns=flexible]
# ac, d |-? q (using previous-results table rt)
def evaluate(ac, rt, q, d)
  return rt[q, d] if rt.contains q, d
  p = cond(ac, rt, q, d)
  return (Proven, rt.update q, d, p) if p.isValid 
  p = canSay_CanActAs(ac, rt, q, d)
  return (Proven, rt.update q, d, p) if p.isValid 
  return (Failure, rt.update q, d, Failure)

# Try can-say and can-act-as
def canSay_CanActAs(ac, rt, q, d)
  ac.constants.each do |c| 
    if c.is_a :subject
      p = canActAs ac, rt, q, d
      return Proven if p.isValid
    elsif c.is_a :speaker
      p = canSay ac, rt, q d
      return Proven if p.isValid
  return Failure
\end{lstlisting}
\end{minipage}
\begin{minipage}[b]{0.5\linewidth}
\begin{lstlisting}[language=Ruby, basicstyle=\ttfamily\scriptsize, keywordstyle=\scriptsize\slshape, columns=flexible]
# Try cond rule
def cond(ac, rt, q, d)
  ac.add q.fetch if q.isFetchable
  ac.assertions.each do |a|
    if (u = q.unify a.consequent) &&
       (a = u.substitution a).variables == none
      return checkConditions ac, rt, a, d
  return Failure

# Evaluate the antecedents
def checkConditions(ac, rt, a, d)
  getVarSubstitutions(a, ac.constants).each do |s|
    sa = s.substitute a
    if sa.antecedents.all 
        { |a| evaluate(ac, rt, a, d).isValid }
      p = evaluateC sa.constraint
      return Proven if p.isValid
  return Failure
\end{lstlisting}
\end{minipage}
\caption{Partial-pseudocode for AppPAL evaluation.}
\label{fig:pseudocode}
\end{figure}

\subsection{Policy Idioms}
\label{ssec:idioms}
